---
title: "hw003"
author: "Coni"
date: "September 29, 2018"
output: github_document
---

# Homework 03: Using dplyr/ggplot2 to manipulate and explore data

Load required libraries

```{r load libraries}
suppressPackageStartupMessages(library(dplyr)) # to supress messages after library calls
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(gapminder))
library(knitr)
```

Tasks (pick 3) - for each make a table and a figure, also make observations on what the figures show and the process

# Max and Min GDP per capita for all continents

The simplest way of finding the min or max is by grouping by continent and then summarizing. Disadvantage: it doesn't keep the name of the country or the year (which are kind of important)

```{r min_gdp table}
min_gdp <- gapminder %>%
  group_by(continent) %>%
  summarize(minGDP = min(gdpPercap) ) %>% 
  arrange(continent) 

knitr::kable(min_gdp)
```

The same can be done for the max

```{r max_gdp table}
max_gdp <- gapminder %>%
  group_by(continent) %>%
  summarize(maxGDP = max(gdpPercap) ) %>% 
  arrange(continent) 

knitr::kable(max_gdp)
```

And we can put both in the same table
```{r min_and_max_gdp table}
min_max <- gapminder%>%
  group_by(continent) %>%
  summarize(minGDP = min(gdpPercap), maxGDP = max(gdpPercap))

knitr::kable(min_max)
  
```

We can visualize this with a plot of maxGDP vs minGDP

```{r min_and_max_gdp plot}
min_max %>% ggplot(aes(x=minGDP, y=maxGDP)) +
  geom_point(aes(color = continent))
```



An alternative, more elaborated way of doing it (repurposing code found [here](http://stat545.com/block010_dplyr-end-single-table.html#use-arrange-to-row-order-data-in-a-principled-way)) is the following. One advantage of this way is that we get to know the country and year associated to the lowest and highest GDP per continent.

```{r min and max gdp with countries}
my_df <- gapminder %>%
  select(continent, country, gdpPercap, year) %>% #select relevant variables
  group_by(continent) %>%
  filter(min_rank(desc(gdpPercap)) < 2 | min_rank(gdpPercap) < 2) %>% 
  # rank by gdp in desc order and ascending order and then apply filter to get only these numbers
  arrange(continent) 

knitr::kable(my_df)
```

## spread of GDP per capita within continents

Let's have a look at the spread in gdp per continent with a boxplot and jitter plot first

```{r gdpPercap per continent}
ggplot(gapminder, aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 0.2) +
  scale_y_log10()
```

We see that Asia is the continent with a higher spread (more unequal between countries, and Oceania seems to be the one with the lower spread (less unequeal)

There are a number of different measures of the spread:

1. Standard deviation - `sd()`: square root of the variance (a measure of the distance to the mean)

2. Interquartile range - `IWR()`: is the difference of its upper and lower quartiles. It is a measure of how far apart the middle portion of data spreads in value (more info [here](http://www.r-tutor.com/elementary-statistics/numerical-measures/interquartile-range))

We can use `summarise()`, `sd()` and `IQR()` to get these values:

```{r gdpPercap sd and IQR per continent table}
gapminder %>%
  group_by(continent) %>%
  summarise(sd = sd(gdpPercap), IQR = IQR(gdpPercap)) %>%
  kable()
  
```

These values somewhat agree with the observation from the graph: we see that asia has the highest sd, but the lowerst corresponds to Africa. On the other hand, the highest IQR is that of Europe, and the lowest Africa's.

We can visualize how the standard deviation has changed for each continent in time with the following plot:

```{r gdpPercap sd trends per continent}
gapminder %>%
  group_by(continent) %>%
  mutate(sd = sd(gdpPercap), IQR = IQR(gdpPercap)) %>%
  ggplot(aes(year, sd)) + geom_point(aes(color= continent))
```


#Trimmed mean of life expentacy for different years

Acording to [wikipedia](https://en.wikipedia.org/wiki/Truncated_mean) "a truncated mean or trimmed mean is a statistical measure of central tendency, much like the mean and median. It involves the calculation of the mean after discarding given parts of a probability distribution or sample at the high and low end, and typically discarding an equal amount of both".

- Compute a trimmed mean of life expectancy for different years. 
```{r 10%trimmed mean lifeExp table}
gapminder %>% 
  group_by(continent, year) %>%
  filter(year == 1952 | year == 2007) %>% #filter by year
  summarize(trim_mean = mean(lifeExp, trim = 0.1)) %>% #compute the mean trimming the top and bottom 10%
  kable()
```

This table allows to compare the trimmed mean life expectancy between continents and between the first and last year of measurement in the database.

Now we can make a plot to visualize the evolution of the trimmed mean (10%) of life expectancy per continent. In this case I added a third variable to see how life exp and population correlate


```{r trim_mean_lifeExp trends per continent}
gapminder %>% 
  group_by(continent, year) %>%
  mutate(trim_mean_lifeExp = mean(lifeExp, trim = 0.1), mean_pop = mean(pop)) %>% 
  ggplot(aes(x = year, y = trim_mean_lifeExp)) +
  geom_point(aes(color = continent, size = mean_pop))
```

